---
title: "Installing a Galaxy server for ChIP-seq analysis"
author: "Denis Puthier & Jacques van Helden"
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 5
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 5
  word_document: default
---

## Downloading Galaxy code


We followed the instructions from the Galaxy Web site: 

- [https://wiki.galaxyproject.org/Admin/GetGalaxy](https://wiki.galaxyproject.org/Admin/GetGalaxy)

```{r eval=FALSE}
## get a git clone of galaxy
git clone https://github.com/galaxyproject/galaxy/
cd galaxy     ## Go th the galaxy directory

## Check out the master branch, recommended for production server
git checkout -b master origin/master  
git pull  ## Just in case, we are already up-to-date
```

## Configure the Galaxy server (and get python modules if required)


We first edit the config file to chooe a specific port for Galaxy

```{r eval=FALSE}
cp config/galaxy.ini.sample config/galaxy.ini
```

We then edit this file by setting the port to 8082, because our 8080 is already used for other purposes. 

We performed the following modifications. 
     
admin_users=admin1@address.fr,admin2@univbazar.fr,admin3@gmail.com
port = 8082  # The port on which to listen.
host = 0.0.0.0 ## To enable access over the network
allow_user_deletion = True



## Configuring the Apache server on RSAT

Activate the Apache module rewrite.load

```{r eval=FALSE}
ln -s /etc/apache2/mods-available/rewrite.load  /etc/apache2/mods-enabled/rewrite.load
```

Create a file /etc/apache2/sites-enabled/galaxy.conf with the following content


    <VirtualHost *:80>
    ServerAdmin webmaster@localhost
    ServerSignature Off
    
    # Config pour galaxy ands http://mydomain.com/galaxy
    RewriteEngine on
    RewriteRule ^/galaxy$ /galaxy/ [R]
    RewriteRule ^/galaxy/static/style/(.*) /home/galaxy/galaxy/static/june_2007_style/blue/$1 [L]
    RewriteRule ^/galaxy/static/scripts/(.*) /home/galaxy/galaxy/static/scripts/packed/$1 [L]
    RewriteRule ^/galaxy/static/(.*) /home/galaxy/galaxy/static/$1 [L]
    RewriteRule ^/galaxy/favicon.ico /home/galaxy/galaxy/static/favicon.ico [L]
    RewriteRule ^/galaxy/robots.txt /home/galaxy/galaxy/static/robots.txt [L]
    RewriteRule ^/galaxy(.*) http://localhost:8082$1 [P]
    #RewriteRule ^/galaxy(.*) http://192.168.1.6:8082$1 [P]
    </VirtualHost>


Restart the Apache server.
```{r eval=FALSE}
sudo service apache2 restart
```

## Starting the galaxy server

```{r eval=FALSE}
sh run.sh 
```

On our internal network, the server becomes available at the address:

[http://192.168.1.6:8082](http://192.168.1.6:8082)


## Registrating 

- open a connection to the Galaxy server
- In the Galaxy menu, run the command  **User -> Register**. Enter the same email address as you declared as admin users. 

## Install Galaxy modules





