---
title: "RNA-seq differential analysis report"
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    theme: cerulean
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
  word_document: default
date: '`r Sys.Date()`'
---

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=FALSE, eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE)
```


```{r Loading R libraries, warning=FALSE}
library(knitr) ## Required to run this Rmd file and produce a markdown file (.md)
library(markdown) ## Required to convert the markdown (.md) file into HTML or pdf documents
library(xlsx) ## To export Excel tables

library("stats4bioinfo") ## Generic library from Jacques van Helden
library("edgeR", warn.conflicts = FALSE, quietly=TRUE) ## Differential expression with RNA-seq
library("DESeq2", warn.conflicts = FALSE, quietly=TRUE, verbose=FALSE) ## Differential expression with RNA-seq
library("limma", warn.conflicts = FALSE, quietly=TRUE) ## Required for vennCounts and vennDiagram
library(gplots, warn.conflicts = FALSE, quietly=TRUE) ## Required for heatmaps.2
library(RColorBrewer, warn.conflicts = FALSE, quietly=TRUE)
library("GenomicFeatures")
```

<!-- !!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
PROJECT-SPECIFIC PARAMETERS: THIS PART SHOULD BE ADAPTER FOR EACH PROJECT -->

## Project information

```{r Required parameters}
if (!exists("parameter.file")) {
  stop("This report requires to specify a variable named 'parameter.file', containing the path to an R file describing the parameters for this analysis.")
}
source(file=parameter.file)
```

<!-- END OF THE PROJECT-SPECIFIC PARAMETERS 
!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!! -->

```{r print project information}

kable(data.frame(project.info))

```


```{r Default parameters}
## In this chunk, we define a set of default parameters for the display and the analysis. 
## These parameters can be modified but it is not necessary to adapt them to each project.

verbosity <- 0

## Color palette for heatmaps. I like this Red-Blue palette because 
## - it suggests a subjective feeling of warm (high correlation)/cold (low correlation)
## - it can be seen by people suffering from red–green color blindness.
cols.heatmap <- rev(colorRampPalette(brewer.pal(9,"RdBu"))(100))

## A trick: to enable log-scaled plots for 0 values, I add an epsilon increment
epsilon <- 0.01
```


```{r Load gene information from GTF file}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

################################################################
## Load gene information from the GTF file 
## (should be the same as used to count tags per gene)

if (exists("gtf.file")) {
  ## Jacques or Lucie, TO DO : check if the method makeTxDbFromGFF allows to get gene names and descriptions.
  ## In principle it should be possible since this info is in the GTF file. 
  ## If not, we migh rewite a parser for GTF. 
  
  verbose(paste("Loading gene annotations from GTF file: ", gtf.file), 2)
  #   library(rtracklayer)  # for the import() function
  #   gr <- import(gtf.file)
  txdb <- makeTxDbFromGFF(file=file.path(dir.base, gtf.file),
                          organism=organism.name,
                          dataSource=gtf.source)
  #seqlevels(txdb)
  #   transcripts <- transcriptsBy(txdb, by="gene")
  #   transcript.table <- as.data.frame(transcripts@unlistData)
  #   cds <- cdsBy(txdb, by="gene")
  #   cds.table <- as.data.frame(cds@unlistData)
  #   exons <- exonsBy(txdb, by="gene")
  #   exon.table <- as.data.frame(exons@unlistData)
  all.genes <- genes(txdb)
  gene.info <- as.data.frame(all.genes)
  gene.info$name <- gene.info$gene_id
  gene.info$entrez.id <- NA
  gene.info$description <- "no description"
} else {
  verbose(paste("No GTF file has been specified"))
  g <- nrow(all.counts)
  gene.info <- data.frame("seqnames"=rep(NA, times=g),
                          "start"=rep(NA, times=g),
                          "end"=rep(NA, times=g),
                          "width"=rep(NA, times=g),
                          "strand"=rep(NA, times=g),
                          "gene_id"=row.names(all.counts),
                          "name"=row.names(all.counts),
                          "entrez.id" = rep(NA, times=g),
                          "description"=rep("no description", times=g))
  row.names(gene.info) <- row.names(all.counts)
}
# View(gene.info)



################################################################
## TEMPORARY: load additional gene information from a tab-delimited file generated with RSAT, 
## because for bacterial genomes there is no obvious way to obtain EntrezIDs.
if (exists("gene.info.file")) {
  verbose(paste("Loading gene information from tab-delimited file", gene.info.file))
  
  gene.info.rsat <- read.delim(file.path(dir.base, gene.info.file), sep="\t", skip=3, header=1)  
  ## A bit tricky: we need to discard rows starting with ";", but we cannot use ";" as comment.char since it is found in descriptions and as separator for names
  comment.lines <- grep(gene.info.rsat[,1], pattern = ";")
  gene.info.rsat <- gene.info.rsat[-comment.lines,]
  row.names(gene.info.rsat) <- gene.info.rsat[,1]
  gene.info.rsat <- gene.info.rsat[row.names(gene.info), ] ## Ensure gene.info.rsat contains same IDs in the same order as gene.info
  
  
  ## Extract additional attributes for the gene.info table
  gene.info$name <- gene.info.rsat[row.names(gene.info), "name"]
  gene.info$entrez.id <- gene.info.rsat[row.names(gene.info), 2]
  gene.info$description <- gene.info.rsat[row.names(gene.info), "description"]
  gene.info$names <- gene.info.rsat[row.names(gene.info), "names"]
  # dim(gene.info.rsat)
  # names(gene.info.rsat)
  # View(gene.info.rsat)
} 
# View(gene.info)

## Export the gene information table to keep a trace of what has been used. 
gene.info.out <- paste(sep="", prefix["general.file"], "_gene_descriptions")
verbose(paste("Exporting gene information table", gene.info.out), 1)
write.table(x = gene.info, row.names = FALSE,
            file = file.path(dir.base, paste(sep=".", gene.info.out, "tab")), sep = "\t", quote=FALSE)
write.xlsx(x = gene.info, row.names = FALSE,
           file = file.path(dir.base, paste(sep=".", gene.info.out, "xlsx")))
verbose(paste(sep="", "\tGene info table\t", gene.info.out), 1)

```


## Parameters

| Parameter | Value |
|-----------|-------|
| Sample descriptions | [`r sample.description.file`](`r file.path(dir.base, sample.description.file)`) |
| Analyses descriptions | [`r design.file`](`r file.path(dir.base, design.file)`) |
| Count table | [`r all.counts.table`](`r file.path(dir.base, all.counts.table)`) |
| Thresholds | E-value $\le$ `r thresholds["evalue"]`; <br>Adjusted p-value (FDR) $\le$ `r thresholds["padj"]`; <br>Fold-change $\ge$ `r thresholds["FC"]`|



## Samples


```{r Read sample description and define condition-specific parameters}
#setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

## Read the sample description file, which indicates the 
## condition associated to each sample ID.
verbose(paste("Reading sample description file: ", sample.description.file), 2)
sample.desc <- read.delim(file.path(dir.base, sample.description.file), sep="\t", 
                          comment=";", header = TRUE, row.names=1)
sample.ids <- row.names(sample.desc)

## Experimental conditions
sample.conditions <- as.vector(sample.desc[,1]) ## Condition associated to each sample
names(sample.conditions) <- sample.ids
# print(sample.conditions)

## Build sample labels by concatenating their ID and condition
sample.desc$label <- paste(sep="_", sample.ids, sample.conditions)

## Define a specific color for each distinct condition
conditions <- unique(sample.conditions) ## Set of distinct conditions
cols.conditions <- brewer.pal(max(3, length(conditions)),"Dark2")[1:length(conditions)]
names(cols.conditions) <- conditions
# print(cols.conditions)

## Define a color per sample according to its condition
sample.desc$color <- cols.conditions[sample.conditions]
# names(cols.samples) <- sample.ids
# print(cols.samples)
```



<!-- TO DO: PRINT TABLE WITH SAMPLE IDS PER CONDITION -->



```{r Loading count table}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

################################################################
## Read the count table
verbose("Loading count table", 2)
all.counts <- read.delim(file.path(dir.base, all.counts.table), row.names=1, sep="\t")
# names(all.counts)
# dim(all.counts)

## Check that the header of all.counts match the sample IDs
ids.not.found <- setdiff(sample.ids, names(all.counts)) ## Identify sample IDs with no column in the count table
if (length(ids.not.found) > 0) {
  stop(paste(length(ids.not.found), "Missing columns in count table", all.counts.table), paste(sep="; ", ids.not.found))
}

################################################################
## Restrict the count table to the sample IDs found in the sample description file
all.counts <- all.counts[, sample.ids]
# names(all.counts)
# dim(all.counts)

########################################################################
## Treatment of 0 values.
## Add an epsilon to 0 values only, in order to enable log-transform and display on logarithmic axes.
verbose(paste("Treating zero-values by adding epsilon =", epsilon), 2)
all.counts.epsilon <- all.counts
all.counts.epsilon[all.counts==0] <- epsilon

## Log-transformed data for some plots. 
all.counts.log10 <- log10(all.counts.epsilon)

```

### Sample-wise statistics

The raw counts table ([`r all.counts.table`](`r file.path(dir.base, all.counts.table)`)) contains `r nrow(all.counts)` rows (features, genes) and `r ncol(all.counts)` columns (conditions, samples).


```{r Sample-wise statistics}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

################################################################
## Compute sample-wise statistics on mapped counts
################################################################
stats.per.sample <- calc.stats.per.sample(sample.desc, all.counts)

################################################################
## Compute the counts per million reads 
################################################################
verbose("Computing CPMs", 1)
## Note: the default normalization criterion (scaling by libbrary sum) 
## is questionable because it is stronly sensitive to outliers 
## (very highly expressed genes).  A more robust normalisation criterion 
## is to use the 75th percentile, or the median. We use the median, somewhat arbitrarily, 
## beause it gives a nice alignment on the boxplots.
cpms.libsum <- cpm(all.counts.epsilon)    ## Counts per million reads, normalised by library sum
cpms.perc75 <- cpm(all.counts.epsilon, lib.size = stats.per.sample$perc75)    ## Counts per million reads, normalised by 75th percentile
cpms.perc95 <- cpm(all.counts.epsilon, lib.size = stats.per.sample$perc95)    ## Counts per million reads, normalised by 95th percentile
cpms.median <- cpm(all.counts.epsilon, lib.size = stats.per.sample$median)    ## Counts per million reads, normalised by sample-wise median count
#cpms <- cpms.median ## Choose one normalization factor for the CPMs used below
cpms <- cpms.perc75 ## Choose one normalization factor for the CPMs used below
cpms.log10 <- log10(cpms) ## Log-10 transformed CPMs, with the epsilon for 0 counts
cpms.log2 <- log2(cpms) ## Log-10 transformed CPMs, with the epsilon for 0 counts

## Compute Trimmed Means of M Values (TMM): TO BE DONE
stats.per.sample$cpm.mean <- apply(cpms, 2, mean)
stats.per.sample$log2.cpm.mean <- apply(cpms.log2, 2, mean)
stats.per.sample$log10.cpm.mean <- apply(cpms.log10, 2, mean)

################################################################
## Export stats per sample
#
# names(stats.per.sample)
# head(stats.per.sample)
verbose("Exporting stats per sample", 1)
sample.summary.file <- paste(sep="", prefix["general.file"], "_summary_per_sample")
write.table(x = stats.per.sample, row.names = TRUE, col.names=NA, 
            file = file.path(dir.base, paste(sep=".", sample.summary.file, "tab")), sep = "\t", quote=FALSE)
write.xlsx(x = stats.per.sample, row.names = TRUE, col.names=NA, 
           file = file.path(dir.base, paste(sep=".", sample.summary.file, "xlsx")))
verbose(paste(sep="", "\tSummary per sample\t", sample.summary.file), 1)
```

```{r print stats per sample}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

## Statistics per sample
stats.per.sample.to.print <- c("Mreads", "sum", "min", "perc25", "mean", "median", "perc75", "max", "fract.below.mean")
kable(stats.per.sample[stats.per.sample.to.print], digits=2,
      caption="Sample-wise statistics")

```


### Sample sizes

```{r library_sizes_barplot, fig.width=6, fig.height=6, fig.cap="Library size barplot. Bars indicate the sum of read counts per sample (library)."}
libsize.barplot(stats.per.sample, plot.file=NULL)
```


### Read count distributions

```{r count_boxplots, fig.width=10, fig.height=10, fig.cap="Read count distributions. Top: raw counts. Bottom: counts per millon reads (CPMs). Left panels: linear scale, which emphasizes  outlier features denoted by very high counts. Rigt panels log counts permit to perceive the distribution of its whole range, including small count values. Null counts are replaced by an epsilon < 1, and appearas negative numbers after log transformation."}
par(mfrow=c(2,2))
  ## Boxplot of raw counts
  count.boxplot(all.counts, stats.per.sample, xlab="Raw counts", 
                main="Raw counts per sample")

  ## Boxplot of log10-transformed counts
  count.boxplot(all.counts.log10, stats.per.sample, xlab="log10(counts)", 
                main="log10(counts) per sample")

  ## Boxplot of CPMs
  count.boxplot(cpms, stats.per.sample, xlab="CPM", 
                main="CPMs per sample")
  
  ## Boxplot of log10-transformed CPMs
  count.boxplot(cpms.log10, stats.per.sample, xlab="log10(CPM)", 
                main="log10(CPMs) per sample")
par(mfrow=c(1,1))
```


### Count correlations between samples

Note: count correlations from raw rads are highly impacted by outlier features. 
<font color="red">TO DO</font>
I should adapt the method to trim the upper and lower count values.

```{r count_correl_heatmap, fig.width=8, fig.height=8}
count.correl.heatmap(all.counts)
```


## Between-replicate reproducibility

Between-replicate count comparisons highlight the general reproducibility of the experiment. In some cases they reveal a strong variability of some particular features, which appear very distant from the diagonal. 

<font color="RED">TO DO</font>
I should develop a method to tag the irreproducible features and/or filter them out from the lists of DEG. 

```{r between-replicate_reproducibility, fig.width=10, fig.height=10, fig.cap="Each plot shows the reproducibility between replicates of a given condition."}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

################################################################
## Analyse between-replicate reproducibility
################################################################
verbose("Plotting betwen-replicate comparisons", 2)
#conditions.with.replicate <- setdiff(conditions, "CI_CI_0H")
cond <- conditions[1] ## Choose first condition for testing without the loop
for (cond in conditions) {
#for (cond in conditions.with.replicate) {
  verbose(paste(sep="", "\tcondition\t", cond), 3)
  
  min.rep.to.plot <- 2 ## Not worth plotting replicates for conditions without replicate
  max.rep.to.plot <- 5 ## Restrict the number of replicates to plot, for the sale of readability
  
  ## Create a specific result directory for this condition
  dir.condition <- file.path(dir.DEG, "per_condition", cond)
  dir.create(path = file.path(dir.base, dir.condition), showWarnings = FALSE, recursive = TRUE)
  
  ## Select the specific data for the current condition (samples, counts, CPMs)
  #all.counts.with.replicate <- setdiff(names(all.counts), "CI_CI_0H_1")
  current.samples <- names(all.counts)[sample.conditions == cond]
  nrep <- length(current.samples)

  current.counts <- all.counts[,current.samples]
  current.counts[current.counts==0] <- epsilon
  current.counts.mean <- apply(current.counts, 1, mean)
  current.counts.var <- apply(current.counts, 1, var)
  current.counts.var[current.counts.var==0] <- min(current.counts.var[current.counts.var>0])/100
  
  current.cpms <- cpms[,current.samples] + epsilon
  current.cpms[current.cpms==0] <- epsilon
  current.cpm.mean <- apply(current.cpms, 1, mean)
  current.cpm.var <- apply(current.cpms, 1, var)
  current.cpm.var[current.cpm.var==0] <- min(current.cpm.var[current.cpm.var>0])/100
  
  ################################################################
  ## Plot pairwise comparisons between replicates (result file can be heavy to open)
   #pdf(file= file.path(dir.condition, paste(sep = "", "between-replicate_counts_plot_", cond, ".pdf")), width=10, height=10)
#   # dim(current.counts[,1:min(max.rep.to.plot, nrep)])
  plot(current.counts[,1:min(max.rep.to.plot, nrep)], 
       log="xy", col=cols.conditions[cond], 
#       panel.first=grid(col="#BBBBBB", lty="solid"), ## this does not work, I should check why
       main=paste(cond, " ; raw counts per replicate (log scale)")
  )
#  silence <- dev.off()
  
#   ################################################################
#   ## Plot mean versus variance of CPMs for the current condition.
#   ## BEWARE! This is the between-replicate variance computed for each gene separately, 
#   ## which is quite different from the smoothed estimate of variance used by 
#   ## DESeq2 or edgeR for the negative binomial.
#   ## 
#   ## This "real" variance plot  can be useful to highlight some genes that show 
#   ## an extremely high variance compared to other genes. 
#   ##
#   ## We draw the plot in linear + log scales. Linear scales better highlight the outliers, 
#   ## log scale better shows the general distribution, which covers several orders of magnitude.
#   log.axes <- "xy"
#   for (log.axes in c("", "xy")) {
# #     pdf(file= file.path(dir.condition, paste(sep = "", "CPM_variance-mean_plot_", cond, "_",  log.axes, ".pdf")), width=10, height=10)
#     plot(current.cpm.mean, 
#          current.cpm.var, 
#          log=log.axes, col=cols.conditions[cond], 
#          panel.first=grid(lty="solid", col="#DDDDDD"),
#          xlab=paste("Mean CPM per gene for condition", cond),
#          ylab=paste("Between replicate CPM variance per gene for condition", cond),
#          main=paste(cond, " ; CPM variance/Mean plot"))
#     abline(a=0,b=1, lty="dashed", col="green", lwd=2) ## Milestone for Poisson distributions: var = mean
#     # silence <- dev.off() 
#   }
}

```


## Detection of differentially expressed genes

### Design

```{r Read the design, warning=FALSE}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

## Read the design file, which indicates the anlayses to be done.
## Each row specifies one differential expression analysis, which 
## consists in comparing two conditions. 
verbose(paste("Reading design file: ", design.file), 2)
design <- read.delim(file.path(dir.base, design.file), sep="\t", 
                     comment=";", header = T, row.names=NULL)
```

<!-- TO DO: print a table with the design -->


```{r differential expression analysis}
# setwd(dir.main) ## !!!!! I don't understand why I have to reset the working directory at each chunk

i <- 1
comparison.summary <- design ## Initialize a summary table for each DEG analysis
comparison.summary$prefixes <- paste(sep="_", design$cond1, "vs", design$cond2)

for (i in 1:nrow(design)) {
  ## Identify samples for the first condition
  cond1 <- as.vector(design[i,1])  ## First condition for the current comparison
  samples1 <- sample.ids[sample.conditions == cond1]
   if (length(samples1) < 2) {
     stop(paste("Cannot perform differential analysis. The count table contains less than 2 samples for condition", cond1))
   }
  
  ## Identify samples for the second condition
  cond2 <- as.vector(design[i,2])  ## Second condition for the current comparison
  samples2 <- sample.ids[sample.conditions == cond2]
   if (length(samples2) < 2) {
     stop(paste("Cannot perform differential analysis. The count table contains less than 2 samples for condition", cond2))
   }
  
  verbose(paste(sep="", "\tDifferential analysis\t", i , "/", nrow(design), "\t", cond1, " vs ", cond2), 1)
  
  ## Create a specific result directory for this differential analysis
  prefix["comparison"] <- comparison.summary$prefixes[i]
  dir.analysis <- file.path(dir.DEG, paste(sep="", prefix["comparison"]))
  comparison.summary$dir.analysis <- dir.analysis
  dir.create(path = file.path(dir.base, dir.analysis), showWarnings = FALSE, recursive = TRUE)
  dir.figures <- file.path(dir.analysis, "figures")
  comparison.summary$dir.figures <- dir.figures
  dir.create(path = file.path(dir.base, dir.figures), showWarnings = FALSE, recursive = TRUE)
  prefix["comparison_file"] <- file.path(dir.analysis, prefix["comparison"])
  prefix["comparison_figure"] <- file.path(
    dir.figures, 
    paste(sep="", prefix["comparison"], "_",  suffix.deg))
  
  ## Select counts for the samples belonging to the two conditions
  current.samples <- c(samples1, samples2)
  current.counts <- all.counts[,current.samples]
  # dim(current.counts)  ## For test
  # names(current.counts)
  
  if (sum(!names(current.counts) %in% sample.ids) > 0) {
    stop("Count table contains column names without ID in sample description file.")
  }
  
  ## Define conditions and labels for the samples of the current analysis
  current.sample.conditions <- sample.conditions[current.samples]
  current.labels <- paste(current.sample.conditions, names(current.counts), sep="_")
  
  result.table <- init.deg.table(cpms) 
# View(result.table)
    
  
  ################################################################
  ## DESeq2 analysis
  ################################################################
  deseq2.result.table <- deseq2.analysis(dir.figures=file.path(dir.base, dir.figures))
  result.table <- cbind(result.table, 
                        "DESeq2" = deseq2.result.table[row.names(result.table),])
  # dim(deseq2.result.table)
  # dim(deseq2.result.table)
  # names(deseq2.result.table)
  # View(deseq2.result.table)
  # View(result.table)
  
  ## Save the completed DESeq2 result table
  deseq2.result.file <- paste(sep = "", prefix["comparison_file"], "DESeq2")
  comparison.summary[i,"deseq2"] <- paste(sep=".", deseq2.result.file, c("tab", "xlsx"))
  write.table(x = deseq2.result.table, row.names = FALSE, 
              file = file.path(dir.base, paste(sep=".", deseq2.result.file, "tab")), sep = "\t", quote=FALSE)
  write.xlsx(x = deseq2.result.table, row.names = FALSE, 
             file = file.path(dir.base, paste(sep=".", deseq2.result.file, "xlsx")))
  verbose(paste(sep="", "\t\tDESeq2 result file\t", deseq2.result.file), 1)
  

  ################################################################
  ## edgeR analysis
  ################################################################
  edger.result.table <- edger.analysis(dir.figures=file.path(dir.base, dir.figures))
  result.table <- cbind(result.table, 
                        "edgeR" = edger.result.table[row.names(result.table),])
 
  ## Export edgeR result table
  edger.result.file <- paste(sep="", prefix["comparison_file"], "edgeR")
  comparison.summary[i,"edger"] <- paste(sep=".", edger.result.file, c("tab", "xlsx"))
  write.table(x = edger.result.table, row.names = FALSE, 
              file = file.path(dir.base, paste(sep=".", edger.result.file, "tab")), sep = "\t", quote=FALSE)
  write.xlsx(x = edger.result.table, row.names = FALSE, 
             file = file.path(dir.base, paste(sep=".", edger.result.file, "xlsx")))
  verbose(paste(sep="", "\t\tedgeR result file\t", edger.result.file), 1)

  ################################################################
  ## Compute main stats from edgeR and DESeq2
  ################################################################
  result.table$mean.FC <- (result.table$edgeR.FC + result.table$DESeq2.FC)/2 ## Arithmetic mean for fold-changes
  result.table$mean.log2FC <- (result.table$edgeR.log2FC + result.table$DESeq2.log2FC)/2 ## Arithmetic mean for log-fold-changes
  result.table$padj.geom.mean <- signif(digits=3, sqrt(result.table$edgeR.padj * result.table$DESeq2.padj)) ## geometric mean for p-values
  result.table[is.na(result.table$padj.geom.mean), "padj.geom.mean"] <- 1
  result.table$mean.evalue <- signif(digits=3, sqrt(result.table$edgeR.evalue * result.table$DESeq2.evalue)) ## Geometric mean for e-value
  result.table$sign <- (result.table$edgeR.sign + result.table$DESeq2.sign)/2
  result.table[is.na(result.table$sign), "sign"] <- 0

  ## Sort the result table according to the mean adjusted p-value
  result.table <- result.table[order(result.table$padj.geom.mean),]
  
  ################################################################
  ## Save result table
  ################################################################
  result.file <- paste(sep = "", 
                       prefix["comparison_file"], 
                       "_", suffix.deg, "_DESeq2_and_edgeR")
  comparison.summary[i,"result.table"] <- paste(sep=".", result.file, c("tab", "xlsx"))
  write.table(x = result.table, row.names = FALSE,
              file = file.path(dir.base, paste(sep=".", result.file, "tab")), sep = "\t", quote=FALSE)
  write.xlsx(x = result.table, row.names = FALSE,
              file = file.path(dir.base, paste(sep=".", result.file, "xlsx")))
  verbose(paste(sep="", "\t\tMerged result file\t", result.file), 1)
  
  
  ################################################
  ## Export tables with only the differentially expressed genes 
  ## for the current analysis, and separate the up- and down-regulated genes
  ## 
  selected.columns <- c("gene_id", "name", "description", 
                        "DESeq2.FC", "edgeR.FC", "mean.FC", "mean.log2FC",
                        "DESeq2.padj", "edgeR.padj", "padj.geom.mean", "mean.evalue")

  up.genes <- result.table$edgeR.DEG & result.table$DESeq2.DEG & (result.table$sign > 0) 
  down.genes <- result.table$edgeR.DEG & result.table$DESeq2.DEG & (result.table$sign < 0) 
  selected.genes <- up.genes + down.genes
  sum(selected.genes)
  comparison.summary[i,"N.down"] <- sum(down.genes)

  verbose("Exporting tables with up-regulated genes only")
  deg.file.up <- paste(sep = "", 
                    prefix["comparison_file"], 
                    "_", suffix.deg, "_selected_DEG_up")
  comparison.summary[i,"deg.table.up"] <- paste(sep=".", deg.file.up, c("tab", "xlsx"))
  write.table(x = result.table[up.genes, selected.columns], row.names = FALSE,
              file = file.path(dir.base, paste(sep=".", deg.file.up, "tab")), sep = "\t", quote=FALSE)
  write.xlsx(x = result.table[up.genes, selected.columns], row.names = FALSE,
              file = file.path(dir.base, paste(sep=".", deg.file.up, "xlsx")))
  verbose(paste(sep="", "\t\tSelected DEG file with up-regulated genes\t", deg.file.up), 1)

  verbose("Exporting tables with down-regulated genes only")
  deg.file.down <- paste(sep = "", 
                    prefix["comparison_file"], 
                    "_", suffix.deg, "_selected_DEG_down")
  comparison.summary[i,"deg.table.down"] <- paste(sep=".", deg.file.down, c("tab", "xlsx"))
  write.table(x = result.table[down.genes, selected.columns], row.names = FALSE,
              file = file.path(dir.base, paste(sep=".", deg.file.down, "tab")), sep = "\t", quote=FALSE)
  write.xlsx(x = result.table[down.genes, selected.columns], row.names = FALSE,
             file = file.path(dir.base, paste(sep=".", deg.file.down, "xlsx")))
  verbose(paste(sep="", "\t\tSelected DEG file with down-regulated genes\t", deg.file.down), 1)

}

## Add a column with links to the result directories
comparison.summary$dir.analysis.link <- paste (
  sep="", "[", comparison.summary$dir.analysis, "]",
  "(", file.path(dir.base, comparison.summary$dir.analysis), ")")
comparison.summary$dir.figures.link <- paste (
  sep="", "[",comparison.summary$dir.figures,"]",
  "(", file.path(dir.base, comparison.summary$dir.figures), ")")
comparison.summary$deg.table.up.link <- paste (
  sep="", "[",comparison.summary$deg.table.up,"]",
  "(", file.path(dir.base, comparison.summary$deg.table.up), ")")
comparison.summary$deg.table.down.link <- paste (
  sep="", "[",comparison.summary$deg.table.down,"]",
  "(", file.path(dir.base, comparison.summary$deg.table.down), ")")
comparison.summary$result.table.link <- paste (
  sep="", "[",comparison.summary$result.table,"]",
  "(", file.path(dir.base, comparison.summary$result.table), ")")
comparison.summary.to.print <- comparison.summary[
  c("cond1","cond2", 
    "dir.analysis.link",
    "dir.figures.link",
    "result.table.link")]

```




```{r print_analysis_summary_table, comment="", eval=FALSE}
library(ascii)
print(ascii(comparison.summary.to.print), type = 'pandoc')
```

```{r, comment="", eval=FALSE}
library(xtable)
print(xtable(comparison.summary.to.print))
```

```{r, comment="", eval=FALSE}
# library(pander)
# pandoc.table(comparison.summary.to.print)
```
```{r eval=TRUE, eval=FALSE}
library(pander)
pander(comparison.summary.to.print, style = 'rmarkdown')
```


* * * 

## Directories and files

### Input files

| Content   | File  |
|-----------|-------|
| Base directory (relative to this report) | [`r dir.base`](`r dir.base`) |
| Raw counts | [`r all.counts.table`](`r file.path(dir.base, all.counts.table)`) |
| Gene annotations (GTF file) | [`r gtf.file`](`r file.path(dir.base, gtf.file)`) |

### General output files

| Content   | File  |
|-----------|-------|
| Output directory | [`r dir.DEG`](`r file.path(dir.base, dir.DEG)`) |
| Gene info table | [`r gene.info.out`](`r file.path(dir.base, gene.info.out)`) |
| Sample summary table | [`r sample.summary.file`](`r file.path(dir.base, sample.summary.file)`) |

### Differential analysis

```{r links_to_results, eval=TRUE, cache=FALSE}
library(knitr)

## Selection of up-regulated genes
kable(comparison.summary[, c("cond1", "cond2", "deg.table.up.link")], 
      caption="Up-regulated genes",
      row.names=FALSE)

## Selection of down-regulated genes
kable(comparison.summary[, c("cond1", "cond2", "deg.table.down.link")], 
      caption="Down-regulated genes",
      row.names=FALSE)

## Full result table (with *many* columns)
kable(comparison.summary[, c("cond1", "cond2", "result.table.link")], 
      caption="Detailed result files",
      row.names=FALSE)

## Output directories per comparison
kable(comparison.summary[, c("cond1", "cond2", "dir.analysis.link", "dir.figures.link")], 
      caption="Output directories",
      row.names=FALSE)

```

```{r job done message}
message("Job done")
```


* * *
